---
layout: post
title: Literate Blogging
tags: [jekyll, git, markdown, r]
---

> Literate programming is a methodology that combines a programming language with a documentation language, thereby making programs more robust, more portable, more easily maintained, and arguably more fun to write than programs that are written only in a high-level language. The main idea is to treat a program as a piece of literature, addressed to human beings rather than to a computer. The program is also viewed as a hypertext document, rather like the World Wide Web.<!--more-->

Literatue Programming by [Donald Knuth](http://www-cs-faculty.stanford.edu/~uno/) [-@knuth_1983] is a seminal book  that introduces the concept of literate programming. Using technology available in 2014 we can make a small but important change to the last sentence:

> The program is also viewed as a hypertext document on the World Wide Web.

This blog post is an example for such a document. The page is written in **markdown** (markdown file available [here](https://github.com/mfenner/mfenner.github.io/blob/source/_posts/2014-04-04-literate-blogging.Rmd)), and all embedded code was executed when this page was generated, i.e. when the markdown was converted to HTML and the blog post was published. To demonstrate this I have embedded code in three different languages below - the output is the second code block.

In R you have

```{r}
cat('Hello, R world!\n')
```

Or Python

```{r engine='python'}
print "Hello, Python world!"
```

Or Ruby

```{r engine='ruby'}
puts 'Hello, Ruby world!'
```

You can also embed code within text blocks (inline), so that `3.48 * 723` becomes **`r 3.48 * 723`**. Another important option is to generate figures using the embedded code, e.g. the following figure taken from a recent publication.

```{r density_plot_example, fig.cap="**Figure 1. Citation counts for PLOS Biology articles published in 2010.** Scopus citation counts plotted as a probability distribution for all 197 *PLOS Biology* research articles published in 2010. Data collected May 20, 2013. Median 19 citations; 10% of papers have at least 50 citations. From Fenner [-@fenner2013]."}
# code for figure 1: density plots for citation counts for PLOS Biology
# articles published in 2010

# load May 20, 2013 ALM report
alm <- read.csv("data/alm_report_plos_biology_2013-05-20.csv", stringsAsFactors = FALSE)

# only look at research articles
alm <- subset(alm, alm$article_type == "Research Article")

# only look at papers published in 2010
alm$publication_date <- as.Date(alm$publication_date)
alm <- subset(alm, alm$publication_date > "2010-01-01" & alm$publication_date <=
    "2010-12-31")

# labels
colnames <- dimnames(alm)[[2]]
plos.color <- "#1ebd21"
plos.source <- "scopus"

plos.xlab <- "Scopus Citations"
plos.ylab <- "Probability"

quantile <- quantile(alm[, plos.source], c(0.1, 0.5, 0.9), na.rm = TRUE)

# plot the chart
opar <- par(mai = c(0.5, 0.75, 0.5, 0.5), omi = c(0.25, 0.1, 0.25, 0.1), mgp = c(3,
    0.5, 0.5), fg = "black", cex.main = 2, cex.lab = 1.5, col = plos.color,
    col.main = plos.color, col.lab = plos.color, xaxs = "i", yaxs = "i")

d <- density(alm[, plos.source], from = 0, to = 100)
d$x <- append(d$x, 0)
d$y <- append(d$y, 0)
plot(d, type = "n", main = NA, xlab = NA, ylab = NA, xlim = c(0, 100), frame.plot = FALSE)
polygon(d, col = plos.color, border = NA)
mtext(plos.xlab, side = 1, col = plos.color, cex = 1.25, outer = TRUE, adj = 1,
    at = 1)
mtext(plos.ylab, side = 2, col = plos.color, cex = 1.25, outer = TRUE, adj = 0,
    at = 1, las = 1)

par(opar)
```

All this functionality is provided by [knitr](http://yihui.name/knitr/), a package for the R statistical programming language. knitr has been around for a while, but integration into the [Jekyll](http://jekyllrb.com/) blogging platform is still fragile. Earlier this week at the [rOpenSci hackathon](https://github.com/ropensci/hackathon) (more on this later) a group of us worked hard to improve this integration. We are still not completely done, but the source code is available [here](https://github.com/ropensci/docs). Most importantly, all the conversion happens on the server, and we are only using freely available tools. I have now enabled this functionality for this blog, so expect more code embedded examples in the future.

## References
